Contracts for unions and intersections are hard for Nickel

### 2021-05-17 paper.tex
More on optimizations

\lstdefinelanguage{nickel}{
    keywords=[1]{
      if,
      then,
      else,
      switch,
    },
    keywords=[2]{
      let,
      rec,
      fun,
      in
    },
    keywords=[3]{
      Num,
      Bool,
      Str,
      List,
      Dyn,
      forall,
    },
    keywordsprefix=\#,
    keywords=[4]{
      true,
      false,
      null
    },
    keywords=[5]{
      doc,
      default,
    },
    sensitive=true, % keywords are case-sensitive
    morecomment=[l]{//}, % l is for line comment
    morecomment=**[is][\color{dark-grey}]{\$}{\$},
    morestring=[b]", % defines that strings are enclosed in double quotes
    morestring=[s]{m\#"}{"\#m},
    moredelim=[is][\color{bright-grey}]{\$}{\$},
    literate=
        *{->}{$\rightarrow$}1
        {=>}{$\Rightarrow$}1
        {>=}{$\geq$}1
        {<=}{$\leq$}1
} %

@@

The good performances of modern programs heavily rely on complex optimizations
performed by the compiler or the interpreter. Even more so for functional
languages, because the functional which model is very different from what actual
computers are good at executing, naive code generation can exhibit unacceptable
slowdowns. Functional programs tend to make heavy use of functions, and function
is a costly operation from the point of view of low-level execution.

One such crucial optimization is inlining. Inlining eliminates a function
application by substituting the function for its definition at compile time (or
before execution, for interpreted language). This is especially important for
short functions that are called repeatedly.

% EXAMPLE INLINING

@@

% EXAMPLE CSE

Beyond CSE, there exists other similar transformations, such as Haskell's
let-floating~\ref{} or loop-invariant code motion, which amount to hoist
invariant out from repeated blocks.

More generally, there is a whole family of opitmizations, such as let-floating,
\begin{lstlisting}[language=caml]
type Ordering = Smaller | Equals | Greater
sort : 'a list
       -> ('a -> 'a -> Ordering)
       -> 'a list
random_list : unit -> int list

sort (random_list ()) (fun x y ->
  if x < y then Smaller
  else if x == y then Equals
  else Greater)
\end{lstlisting}

Common subexpression elemination (CSE) etc.

\begin{lstlisting}[language=caml]

\end{lstlisting}

>>>>

\lstdefinelanguage{nickel}{
    keywords=[1]{
      if,
      then,
      else,
      switch,
    },
    keywords=[2]{
      let,
      rec,
      fun,
      in
    },
    keywords=[3]{
      Num,
      Bool,
      Str,
      List,
      Dyn,
      forall,
    },
    keywordsprefix=\#,
    keywords=[4]{
      true,
      false,
      null
    },
    keywords=[5]{
      doc,
      default,
    },
    sensitive=true, % keywords are case-sensitive
    morecomment=[l]{//}, % l is for line comment
    morecomment=**[is][\color{gray}]{\$}{\$},
    morestring=[b]", % defines that strings are enclosed in double quotes
    morestring=[s]{m\#"}{"\#m},
    moredelim=[is][\color{gray}]{\$}{\$},
    literate=
        *{->}{$\rightarrow$}1
        {=>}{$\Rightarrow$}1
        {>=}{$\geq$}1
        {<=}{$\leq$}1
} %

@@

The performance of modern programs heavily relies on non-trivial optimizations
performed by the compiler or the interpreter. Even more so for functional
languages, whose operational semantics is usually very different from the actual
primitives of modern processors: naive code generation can exhibit unacceptable
slowdowns. Functional programs tend to make heavy use of functions, and a function
call is a costly operation from the point of view of low-level execution.

One such crucial optimization is inlining. Inlining eliminates a function
application by directly substituting the function for its definition at compile
time (or before execution, for interpreted language). This is especially
important for small functions that are called repeatedly.

\begin{lstlisting}[language=nickel]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

let subList = fun l1 l2 =>
  elem (lists.head l1) l2
  && subList (list.tail l1) l2
\end{lstlisting}

@@

\begin{lstlisting}[language=nickel]
let elemAtOrLast = fun index list =>
  if index == lists.length list then
    lists.elemAt (lists.length list - 1) list
  else
    lists.elemAt index list
\end{lstlisting}

Similar transformations exist beyond CSE, such as loop-invariant code motion or
some of Haskell's let-floating~\ref{}, which amount to hoist invariant
expressions out from repeated blocks in order to avoid recomputing them. In
particular, take the following example:

\begin{lstlisting}[language=nickel]
let f = fun x =>
  g y (x + y)

// We want
let f =
  let _g = g y in
  fun x => \_g (x + y)
\end{lstlisting}

The soundness of these optimizations is tied the invariance of the semantics of
programs with respect to substitution. Inlining requires that one can replace
the application of a function by its body: this is $\beta$-reduction, and it is
hard to think of a language when it is not a valid transformations. On the other
hand, CSE-like transformations require an expansion property, of the form

$$M[N/x] \simeq let~x~=~M~in~M[x]$$

This does not necessarily holds, and is in particular trivially violated in
presence of side effects:

$$let~f~x~=~print~"hi";~(x+1)~in~(f~1, f~1) \\
\not \simeq let~f~x~=~print~"hi";~(x+1)~in~let~y~=~f~1~in~(y,y)$$

However, this property does hold for \emph{pure} terms, that are terms without
side-effects. Pure languages, or languages with a form of effect tracking such
as Haskell, are able to identify such pure terms. Other languages rely on
sub-approximations (for instance, applying optimizations to arithmetic
expressions).
\newpage
